\documentclass{scrartcl}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{beramono}
\usepackage{enumitem}
\usepackage{etoolbox}
\usepackage{float}
\usepackage{ebgaramond}
\usepackage{listings}
\usepackage{mathpartir}
\usepackage{microtype}
\usepackage{natbib}
\usepackage{titling}
\usepackage{subcaption}
\usepackage{xcolor}
\usepackage{xspace}
\usepackage{hyperref}

\input{tex/macros}
\input{tex/listings}

\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}

\raggedbottom
\setlength{\droptitle}{-8ex}
\setlist[itemize]{leftmargin=*, nosep, topsep=2mm}
\setlist[enumerate]{nosep}

\title{\textbf{\PurePy: A Functional Subset of Python}}
\version{0}{9}{0}

\author{
   Nikolaus Huber\textsuperscript{1} \and
   Dominic Orchard\textsuperscript{2,1} \and
   Jacob Pake\textsuperscript{2} \and
   Roly Perera\textsuperscript{1,3}
}
\date{}

\newcommand{\affiliations}{%
    \textsuperscript{1}\textit{Department of Computer Science and Technology, University of Cambridge, UK} \\
    \textsuperscript{2}\textit{School of Computing, University of Kent, UK} \\
    \textsuperscript{3}\textit{School of Computer Science, University of Bristol, UK}
}

% Override maketitle to avoid internal tabular conflicts
\makeatletter
\renewcommand{\maketitle}{%
    \vspace*{\droptitle}
    \begin{flushleft}
        {\LARGE\bfseries\thetitle\par}
        \ifx\versionstring\@empty
        \else
            \vspace{0.5ex}
            {\Large\normalfont version \versionstring\par}
        \fi
        \vspace{1ex}
        {\Large\theauthor\par}
        \vspace{1ex}
        {\affiliations\par}
    \end{flushleft}
    \vspace{2ex}
}
\makeatother

\begin{document}

\pagestyle{plain}

% Otherwise custom \maketitle allows floats to appear above title
\makeatletter
\let\orig@maketitle\maketitle
\renewcommand{\maketitle}{%
  \orig@maketitle
  \global\@topnum=0
}
\makeatother

\maketitle

\begin{abstract}
\noindent \PurePy is a pure functional subset of Python, providing a common functional language for use by
researchers in programming languages and pedagogy, while providing a front-end familiar and accessible to the
Python community. This specification defines a versioned standard with formal syntax, well-formedness rules
based on a definite assignment analysis, and an operational semantics intended to preserve the meaning of
well-formed \PurePy programs in Python. By combining functional programming with Python-compatible surface
syntax, \PurePy aims to support portable implementations and lower-friction adoption of new language ideas for
modelling, data processing, analysis, and visualisation.
\end{abstract}

\section{Introduction}

\begin{itemize}
\item Define subset of Python called \PurePy, defined as a syntactic subset further restricted by a
well-formedness predicate
\item Define a semantics for \PurePy such that the semantics of a \PurePy program is identical to its
semantics in Python
\item Some well-formedness and evaluation rules could be given by a desugaring, but for now we avoid having
two separate syntaxes
\end{itemize}

Python interoperability is a key goal, but \PurePy is also motivated by the desire to carve out a well-behaved
discipline for scientific programming \emph{within} Python: a sublanguage in which permissive Python idioms
that invite confusion are tightened or excluded.

{\small
\begin{center}
\begin{minipage}[t]{0.4\textwidth}
\vspace{0pt}
\begin{tabular}{ll}
\textbf{To do} & \\
Reserved words & \issue{9} \\
Dataclasses & \issue{8} \\
Set and dict comprehensions & \issue{52} \\
Set and dict literals & \issue{87} \\
\pyinline{import} & \issue{53} \\
Membership operator (\pyinline{in}) & \issue{80} \\
Identity operator (\pyinline{is}) & \issue{81} \\
Chained comparison & \issue{82} \\
\pyinline{match} statement & \issue{83} \\
Enums & \issue{86} \\
\end{tabular}
\end{minipage}\hspace{2em}
\begin{minipage}[t]{0.4\textwidth}
\vspace{0pt}
\begin{tabular}{ll}
\textbf{Under consideration} & \\
Destructuring assignment & \issue{54} \\
f-strings & \issue{55} \\
Default arguments & \issue{56} \\
\pyinline{*args} and \pyinline{**kwargs} & \issue{57} \\
Decorators & \issue{58} \\
Slicing & \issue{59} \\
Star patterns & \issue{84} \\
Or patterns & \issue{85} \\
\end{tabular}
\end{minipage}
\end{center}}

\section{Syntax}

The syntax of the language is given in \figref{syntax}. A \emph{block} is a non-empty sequence of
statements; either a top-level statement, a \pyinline{def} body, or a branch of a conditional. A
\emph{mutual region} $d_1 \ldots d_{n+1}$ is a maximal contiguous non-empty sequence of \pyinline{def}
statements in a block: any non-\pyinline{def} statement in that block, by definition, separates mutual
regions. Maximality is not enforced by the grammar in \figref{syntax}; the parser is responsible for
grouping consecutive \pyinline{def}s into a single region. The well-formedness rules treat each region as
one mutually-recursive binding (see \S\ref{sec:mutual-recursion}).

\subsection{Contexts}

The grammar of contexts and result types is given in \figref{contexts}. Contexts $\Gamma, \Delta$ are partial functions from sets of variables to a \emph{definite assignment} status
in $\B = \set{\statusDefAssigned, \statusNotDefAssigned}$, with $\statusDefAssigned$ indicating
\emph{definitely assigned} and $\statusNotDefAssigned$ indicating \emph{not definitely assigned}. The definite
assignment terminology is borrowed from Java~\cite[\S16]{gosling2025}. (Or think of $\statusDefAssigned$ and
$\statusNotDefAssigned$ as degenerate types ``any'' and ``none''.)  It is useful to represent contexts as
total functions from the set of all variables to $\B_{\bot} = \B \cup \set{\bot}$, with $\bot$ indicating
\emph{undefined}. We write $\Gamma, x : a$ for the context that sends $x$ to $a$ and otherwise agrees with
$\Gamma$, define $\dom(\Gamma) = \set{x \mid \Gamma(x) \neq \bot}$, and write $\varnothing$ for the context
where all variables are undefined. For a set of variables $X$, write $\statusDefAssigned(X)$ for the context
$\set{x : \statusDefAssigned \mid x \in X}$; similarly $\statusNotDefAssigned(X)$.

\input{fig/contexts}

\paragraph{Context operations.}

Contexts admit a non-commutative monoid of sequential composition $\Gamma \runion \Delta$ (``override'') and
an idempotent commutative semigroup of parallel composition $\merge$ (``merge''). Together these form a
\emph{right near-semiring}~\cite{vanhoorn1967}: $\runion$ right-distributes over $\merge$ (Lemma~1), but left
distributivity does not hold. Both operations are defined pointwise via operations on $\B_{\bot}$:
%
\begin{definition}[Override and merge]
For $a, a' \in \B$:
{\small
\begin{align*}
\bot \runion a = a \runion \bot &= a & \bot \merge \bot &= \bot \\
a \runion a' &= a' & a \merge \bot = \bot \merge a &= \statusNotDefAssigned \\
& & a \merge a' &= a \meet a'
\end{align*}}
\end{definition}

\noindent The $\runion$ operation is right-biased and has $\bot$ as left and right unit. The $\merge$
operation lacks a unit; when exactly one argument is defined, the result is $\statusNotDefAssigned$; when
both are defined, it returns the conjunction of their values in $\B$.

\begin{lemma}[Right distributivity]
\[
(\Delta \merge \Delta') \runion \Gamma = (\Delta \runion \Gamma) \merge (\Delta' \runion \Gamma).
\]
\end{lemma}

\noindent Left distributivity ($\Gamma \runion (\Delta \merge \Delta') = (\Gamma \runion \Delta) \merge
(\Gamma \runion \Delta')$) does not hold. If it did, the following two programs would be equivalent, since
sequential composition with a preceding assignment could be distributed into the branches of a conditional.
But they are not: the left program is ill-formed, because a variable assigned before a conditional does not
remain definitely assigned if only some branches assign it (see also \S\ref{sec:local-assignment}).
\vspace{-\medskipamount}

\begin{minipage}[t]{0.48\textwidth}
\begin{python}
def f(b):
    x = 1
    if b:
        x = 2
    else:
        pass
    return x  # error
\end{python}
\end{minipage}
\hfill
\begin{minipage}[t]{0.48\textwidth}
\begin{python}
def f(b):
    if b:
        x = 1
        x = 2
    else:
        x = 1
        pass
    return x  # ok
\end{python}
\end{minipage}

Context operations $\runion$ and $\merge$ lift to a right semiring over result types $R$, with
$\Assigns{\varnothing}$ as left and right unit for $\runion$ and $\Returns$ as unit for $\merge$ and zero for
$\runion$.

\begin{definition}[Override and merge on result types]
{\small
\begin{align*}
\Assigns{\Delta} \merge \Assigns{\Delta'} & = \Assigns{(\Delta \merge \Delta')} \\
R \merge \Returns = \Returns \merge R & = R \\
\\
\Assigns{\Delta} \runion \Assigns{\Delta'} & = \Assigns{(\Delta \runion \Delta')} \\
R \runion \Returns = \Returns \runion R & = \Returns
\end{align*}}
\end{definition}

\noindent The $\Returns \runion R$ case never arises because the well-formedness judgement prohibits
unreachable statements (\S\ref{sec:unreachable}).

\begin{lemma}[Right distributivity]
\[
(R \merge R') \runion S = (R \runion S) \merge (R' \runion S)
\]
\end{lemma}

\input{fig/syntax}

\paragraph{Operators.}

The metavariable $\odot$ ranges over the supported binary operators and $\ominus$ over the supported unary
operators, both enumerated in \figref{precedence}. The grammar of \figref{syntax} is intentionally ambiguous
on operator forms; disambiguation is by the precedence and associativity given in \figref{precedence},
which follow Python.

\input{fig/precedence}

\subsection{Modules}

A \PurePy module $m$ is the contents of a single source file (\figref{syntax}). Well-formedness of $m$ is
checked under the initial context $\Gamma_m$, which binds at status $\statusDefAssigned$ the names of
\figref{predefined}: the names of the predefined module \pyinline{builtins}; and the variable
\pyinline{__name__}, set per module to
\pyinline{"__main__"} in the entry-point module and the module's own name otherwise.

\begin{figure}[H]
\small
\centering
\begin{tabular}{lp{0.6\textwidth}}
\hline
\textbf{Source} & \textbf{Names} \\
\hline
\pyinline{builtins} module & \pyinline{print}, \pyinline{len}, \pyinline{range}, \pyinline{type}, \pyinline{True}, \pyinline{False}, \pyinline{None} \\
Module-local & \pyinline{__name__} \\
\hline
\end{tabular}
\caption{Predefined names available in every \PurePy module.}
\label{fig:predefined}
\end{figure}

\subsection{Well-formedness}

\paragraph{Auxiliary metafunctions.}

Three syntactic analyses support the well-formedness rules: $\assignsS(s)$, the set of variables assigned
anywhere in $s$ (without descending into nested function definitions, which have their own scope, so that
$\assignsS(s) = \dom(\Delta)$ when $s : \Assigns{\Delta}$); $\binds(p)$, the set of variables introduced by a
pattern $p$; and $\capturesS(s)$, the set of variables from the enclosing scope that are captured by closures
(lambdas or nested function definitions) within $s$, with companion $\capturesE(e)$ on expressions. All are
defined in \figref{aux}. The functions $\assignsS$ and $\capturesS$ extend pointwise to blocks via
$\assignsS(s \cons b) = \assignsS(s) \cup \assignsS(b)$ (similarly $\capturesS$), with the singleton block $s$
identifying with the underlying statement.

\input{fig/aux}

\paragraph{Pattern subsumption.}

The preorder $p \subsumed p'$, defined in \figref{subsumption}, holds when every value matched by $p$ is also
matched by $p'$. It is used by the well-formedness rule for $\pyinline{match}$ to rule out unreachable cases,
and to check exhaustiveness: with the current pattern set, this reduces to requiring a catch-all (var or
wildcard) case.

\input{fig/subsumption}

\paragraph{Well-formed pattern lists.}

A pattern list is well-formed (\figref{well-formed-pat}) when each pattern is linear (no repeated variable
names) and no earlier pattern subsumes a later one. The \pyinline{match} well-formedness rules use this
judgement to enforce reachability and linearity of case patterns.

\input{fig/well-formed-pat}

\paragraph{Well-formedness judgement.}

Every well-formed statement either \emph{definitely returns}, or \emph{definitely assigns} a context
$\Delta$, which may be empty. The base cases are that \pyinline{return e} definitely returns, and \pyinline{x
= e} definitely assigns $\set{\pyinline{x} : \statusDefAssigned}$. Block well-formedness is given in
\figref{well-formed}; expression well-formedness in \figref{well-formed-expr}.

\input{fig/well-formed}
\input{fig/well-formed-expr}

\subsection{Examples}

\subsubsection{Assignment as definition}

Although \PurePy is pure and all variable usage is well-scoped, local variables are introduced using an
\emph{assignment statement}, for compatibility with Python. If a variable is definitely assigned in any branch
of a conditional, then for that variable to be usable after the conditional, every other branch of the
conditional must either definitely return, or definitely assign that variable too. Thus the following are both
well-formed:

\begin{figure}[H]
\centering
\begin{subfigure}{0.48\textwidth}
\centering
\begin{python}
if a < b:
   comp = "smaller"
else:
   comp = "bigger"
return comp
\end{python}
\end{subfigure}
\hfill
\begin{subfigure}{0.48\textwidth}
\centering
\begin{python}
if a < b:
   return "smaller"
else:
   comp = "bigger"
return comp
\end{python}
\end{subfigure}
\end{figure}

\noindent The definite assignment set for a conditional that has at least one branch which definitely assigns
is calculated by merging the definite assignment sets for all branches which definitely assign, using the
$\merge$ operator on environments.

\subsubsection{Local assignment in conditional clauses}
\label{sec:local-assignment}

For flexibility, we permit \emph{branch-local} variables. The assignment of \pyinline{x} below (left) is
permitted, and can be used locally in that branch of the conditional:

\begin{figure}[H]
\centering
\begin{subfigure}{0.48\textwidth}
\centering
\begin{python}
def foo():
  z = 6
  if b:
    x = 3
    y = x + 1
  else:
    y = 0
  return z
\end{python}
\caption{Branch-local definition of \pyinline{x}}
\end{subfigure}
\hfill
\begin{subfigure}{0.48\textwidth}
\centering
\begin{python}
def foo():
  x = 6
  if b:
    x = 3
    y = x + 1
  else:
    y = 0
  return x # error: partially defined
\end{python}
\caption{Now with use of \pyinline{x} after the conditional}
\end{subfigure}
\end{figure}

\noindent However such variables only appear to be branch-local. The definition of $\merge$ actually ensures
that a variable assigned in any branch is considered assigned by the conditional; however unless it is
assigned in \emph{every} branch that does not definitely return, it will be bound to $\statusNotDefAssigned$ in the
corresponding environment. Any attempt to refer to such a variable is illegal, as it refers to a variable
which may not be assigned. (Thus the right-hand program is ill-formed, despite the enclosing \pyinline{x = 6},
because in the \pyinline{return x} statement, the inner assignment shadows the the outer one, and the inner
assignment is only partial.) If we were to allow the program on the right, there would be no straightforward
way to assign it a semantics which is both pure, and in agreement with Python, since in Python the function
returns \pyinline{3} in the case where \pyinline{b} is \pyinline{True}.

\subsubsection{Function-level scoping of local variables}

In Python, all local variables are scoped to the entire function body, not just from the point of
assignment onwards. The $\assignsS$ function in the \ruleName{mutual} rule reflects this: all variables
assigned anywhere in the function body are introduced at $\statusNotDefAssigned$ before the body is checked. This means
the following program is statically rejected by \PurePy, whereas Python rejects it at runtime with an
\pyinline{UnboundLocalError}:

\begin{python}
y = 7

def f():
    x = y  # error: y is local (due to assignment below)
    y = 8

f()
\end{python}

\noindent Because \pyinline{y} appears on the left-hand side of an assignment in \pyinline{f}, it is in
$\assignsS(b)$ for the body of \pyinline{f}, and is therefore introduced at $\statusNotDefAssigned$ at the top of the
function. The reference to \pyinline{y} in \pyinline{x = y} then fails the well-formedness check, since
$\Gamma(\pyinline{y}) = \statusNotDefAssigned$ rather than $\Gamma(\pyinline{y}) = \statusDefAssigned$.

\subsubsection{Reassignment of captured variables}
\label{sec:captured-vars}

Once a variable has been captured by a closure, it cannot be reassigned in the enclosing scope, because
its Python semantics would depend on mutation. Consider the following program:

\begin{python}
def f():
    x = 5
    def g():
        return x
    x = 6
    return x + g()
\end{python}

\noindent In Python, \pyinline{g} closes over the \emph{mutable binding} of \pyinline{x}, so
\pyinline{g()} returns \pyinline{6} and the result is \pyinline{12}. In a pure semantics where
\pyinline{x = 6} shadows the earlier binding, \pyinline{g()} would return \pyinline{5} and the result
would be \pyinline{11}. To preserve agreement with a subset of Python, we exclude this program from
\PurePy: the side condition in the \ruleName{seq} rule requires that no variable captured by
$s$ is reassigned in the rest of the block $b$. Here \pyinline{x} is captured by
the definition of \pyinline{g} and reassigned by \pyinline{x = 6}, so the program is rejected. This is
analogous to Java's requirement that local variables captured by a lambda must be \emph{effectively
final}~\cite[\S15.27.2]{gosling2025}: local variables are allocated on the stack, and so only the value
of a local variable can be captured by a closure.

The same issue arises when a variable is captured and reassigned in the \emph{same} statement:

\begin{python}
x = 5
def x():
    return x
x()  # Python: returns the function (late binding); pure: returns 5
\end{python}

\noindent The side condition on the \ruleName{assign} rule rejects programs like this, requiring that the
variable being bound is not captured by the expression. The \ruleName{mutual} rule, by contrast, binds
each $f_i$ in the typing context for every body $b_i$, so the function names may legitimately appear in
$\capturesS$ of those bodies. This is what permits self-recursion (the singleton-region case) and
mutual recursion within a region.

\subsubsection{Mutual recursion}
\label{sec:mutual-recursion}

The \ruleName{mutual} rule binds the names $f_1, \ldots, f_{n+1}$ of a mutual region $d_1 \ldots d_{n+1}$
simultaneously. Each body $b_i$ is checked in a context that includes all the $f_j$ as definitely
assigned, so any $f_j$ may reference any other $f_k$ within the region (the singleton case $n = 0$
permits self-recursion). The names in a region must be distinct.

Inserting any non-\pyinline{def} statement between two \pyinline{def}s splits them into separate regions,
breaking the mutual reference:

\begin{python}
def even(n):
    if n == 0:
        return True
    return odd(n - 1)

x = even(0)  # intervening statement separates the regions

def odd(n):
    if n == 0:
        return False
    return even(n - 1)
\end{python}

\noindent This program is rejected statically: \pyinline{even} forms a singleton region in which
\pyinline{odd} is not yet bound, so the reference to \pyinline{odd} in its body fails the
\ruleName{var} rule. Python accepts it: the call \pyinline{even(0)} does not actually reach
\pyinline{odd}, and Python's late binding means \pyinline{odd} is only looked up when needed.

Each branch of a conditional may contain its own mutual regions, subject to the usual definite assignment
rules. The following is well-formed because both branches define both names:

\begin{python}
b = True

if b:
    def f():
        return g()
    def g():
        return "via mutual region"
else:
    def f():
        return "f independent"
    def g():
        return "g independent"

print(f())
print(g())
\end{python}

A region containing two definitions of the same name is rejected directly by the \ruleName{mutual}
rule's distinct-names side condition:

\begin{python}
def g():
    return 0

def f():
    return g()
def g():     # error: g defined twice in the same region
    return 1
\end{python}

\noindent (Python accepts this; late binding makes \pyinline{f}'s call to \pyinline{g} resolve to the
second \pyinline{def g}.)

Wrapping the second \pyinline{def f; def g} pair in a conditional sidesteps the duplicate-name issue:
the outer \pyinline{def g} and the inner \pyinline{def g} are now in different blocks, so they no longer
fall in the same region. One might still expect a captured-variable problem, since the inner
\pyinline{def g} could be seen as rebinding the outer \pyinline{g} that \pyinline{f} captures. No such
problem arises:

\begin{python}
def g():
    return 0

if b:
    def f():
        return g()
    def g():
        return 1
\end{python}

\noindent The inner \pyinline{def f; def g} form a branch-local mutual region, so \pyinline{f}'s body
resolves \pyinline{g} to its sibling rather than to the outer \pyinline{g}. The two are bound
simultaneously, so \pyinline{f} cannot be called before the inner \pyinline{g} exists, and the value
\pyinline{f} sees is the same as the value Python's late binding finds at call time.

\subsubsection{Pattern variables escape the match}

A variable bound by a \pyinline{case} pattern is in scope after the \pyinline{match} statement, just as
a variable assigned inside a conditional branch is in scope after the conditional. Python treats matching
this way; \PurePy follows.

\begin{python}
v = 5
match v:
    case x:
        pass
print(x)    # prints 5
\end{python}

Here the catch-all \pyinline{case x:} binds \pyinline{x} in the body's scope, and the binding survives into
the enclosing scope. Definite assignment after the match follows the same merge logic as a conditional: if
every case binds the variable (or returns), it is definitely assigned afterwards; otherwise it is not (see
also \S\ref{sec:local-assignment}).

\subsubsection{Unreachable code}
\label{sec:unreachable}
A non-empty sequence of statements returns only if the last statement returns, and all preceding statements
assign. Thus the following is illegal:

\begin{python}
def foo(x):
  return x + 1
  return x    # error: unreachable
\end{python}

\subsubsection{Conditionals without \pyinline{else} clauses}

An \pyinline{if} statement without an \pyinline{else} clause is equivalent to an \pyinline{if} statement with
an \pyinline{else} clause whose body is a $\exPass$ statement; such a statement never returns.

\subsubsection{Implicit \pyinline{None}}

A \pyinline{return} statement without an expression is equivalent to \pyinline{return None}.

\subsubsection{Implicit \pyinline{return}}

A function whose body $s$ assigns rather than returns is equivalent to a function whose body is $s$ followed
by \pyinline{return None}. This ensures every function returns, even if it is not explicitly specified.

\subsubsection{\pyinline{assert} without a message}

An \pyinline{assert e} statement without a message is equivalent to \pyinline{assert e, ""}.

\subsection{Unsupported Python features}

The following Python features are excluded from \PurePy as they are incompatible with purity or
functional programming:
\begin{itemize}
\item \pyinline{while} loops, \pyinline{for} loops, \pyinline{break}, \pyinline{continue}
\item \pyinline{async}/\pyinline{await}
\item \pyinline{global}, \pyinline{nonlocal}
\item \pyinline{del}
\item \pyinline{try}/\pyinline{except}/\pyinline{raise}
\item Augmented assignment (\pyinline{+=}, \pyinline{-=}, etc.)
\item Walrus operator (\pyinline{:=})
\item \pyinline{yield}/\pyinline{yield from}
\item \pyinline{class} definitions (dataclasses may be added; see \S2 to-do list)
\item \pyinline{with} statements
\item List patterns matching tuple values, or tuple patterns matching list values. In \PurePy a list pattern matches only a list, and a tuple pattern matches only a tuple. Python permits either pattern against either kind of sequence.
\end{itemize}

\section{Operational semantics}

For a statement $s$, the result $\assigns{\rho}$ represents \emph{normal completion} of $s$, with accumulated
environment $\rho$; the result $\returns{v}$ represents \emph{abrupt completion} of $s$ with return value $v$.
Composing statements sequentially composes their results according to the following partial monoid, analogous
to sequential composition of result types:
{\small
\begin{align*}
\assigns{\rho} \runion \assigns{\rho'} & = \assigns{(\rho \runion \rho')} \\
\assigns{\rho} \runion \returns{v} & = \returns{v}
\end{align*}}

\noindent with $\assigns{\varnothing}$ as left and right unit. Function definitions delimit the scope of these
effects, so as black boxes functions are always pure.

\input{fig/eval}

\noindent The pattern-matching judgement $p, v \matches \rho$ is a deterministic partial function on $(p,
v)$. We write $p, v \notmatches$ (with no right-hand side) for $\nexists \rho.\, p, v \matches \rho$, i.e.\
matching is undefined.

\input{fig/match}

\input{tex/related-work}

\bibliographystyle{plain}
\bibliography{tex/refs,tex/additional-refs}

\ifdefined\openterms
\input{tex/appendix-open-terms}
\fi

\end{document}
